Individual and Differential Harm in Redistricting

### 2024-12-04 paper/harm.qmd
rerender

---
title: Individual and Differential Harm in Redistricting
authors:
  - name: Anonymized Author(s)
    affiliation: Anonymized Institution
abstract: |
  Social scientists have developed dozens of measures for assessing partisan bias in redistricting.
  But these measures are not easily adapted to other groups, 
  including groups defined by race, class, or geography.
  Nor are they applicable to single- or no-party contexts, such as local redistricting.
  To overcome these limitations, we propose a unified framework of harm for evaluating 
  the impacts of a districting plan on individual voters and the groups to which they belong. 
  We consider a voter harmed if their chosen candidate is not elected under the current plan, 
  but would be under a different plan.
  Harm improves on existing measures by both focusing on the choices of individual voters 
  and directly incorporating counterfactual plans.
  We discuss strategies for estimating harm, and demonstrate the utility of our framework 
  through analyses of partisan gerrymandering in New Jersey, voting rights 
  litigation in Alabama, and racial dynamics of Boston City Council elections.
keywords:
  - representation
  - redistricting
  - voting rights
  - individual harm
bibliography: references.bib
biblio-style: apsr
output:
  rticles::arxiv_article:
    keep_tex: true
    includes:
      in_header: "header.tex"
    latex_engine: pdflatex
fontsize: 12pt
editor_options:
  markdown:
    wrap: sentence
---

@@

The harm framework combines two key ideas, which are motivated at greater length in Section \ref{sec:motivation}.
First, it operates at the level of the individual voter and examines whether each voter is represented by the candidate of their choice.
Doing so both unifies the approach to gerrymandering across contexts and also reflects a core goal of representative democracy: that elected officials represent the preferences of the voting public to the greatest extent possible [see also @brunell2010].

@@

The combination of individual voter choices and counterfactual districting plans defines our notion of *individual harm*: informally, a voter is harmed if they are more likely to be represented by the candidate of their choice under a different districting plan.
And by analyzing how harms fall differentially along partisan, racial, ideological, and other lines, we can connect individual harm to traditional notions of gerrymandering.
Section \ref{sec:motivation} illustrates this with a simple example, and the general definitions are formalized in Section \ref{sec:indiv-harm}.

A key feature of the proposed framework is its clear separation of theoretical quantities of interest and empirical strategies for estimating them [@katz2022]. This distinction is also missed by many critics of the efficiency gap [@veomett2018efficiency; @deford2020symmetry]. The challenge of estimating harm and differential harm from observed data is discussed in Section \ref{sec:estimating}. Recent advances in redistricting simulation, which can generate a large number of random districting plans that respect relevant legal constraints, are critical to this effort [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020].
These algorithmically-generated sets of districting plans can provide a realistic set of counterfactuals for our measures.

The harm framework has important similarities with the common-law approach to voting rights litigation.
To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker; @issakarl98]. This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidate [@2004vieth; @2018league; @2019lewis].
Indeed, a primary factor in the Supreme Court's ruling that partisan gerrymandering is not justiciable was that existing redistricting measures are not clearly tied to individual injury [@2017gill].
A more detailed discussion of the connections between the proposed framework and existing practices for analyzing racial and partisan gerrymandering may be found in Section \ref{sec:connections}.

Harm and differential harm are more than a useful conceptual unification---they are practically useful in a wide variety of redistricting analyses.
A brief study of partisan gerrymandering of New Jersey congressional districts in Section \ref{sec:nj} shows how existing gerrymandering metrics can make incorrect conclusions about the magnitude and even the direction of partisan gerrymanders.
Section \ref{sec:alabama}, studies voting rights litigation for Alabama congressional districts, and uses differential harm to separate the racial and partisan effects of the enacted districting plan.
Section \ref{sec:boston} demonstrates the application of our framework to local redistricting, and quantifies the benefits of race-aware city council district boundaries to Black voters in the city of Boston.
Section \ref{sec:discussion} concludes.

# Motivation {#sec:motivation}

The core idea motivating individual and differential harm is a simple one.
Consider *Minissouri*, a stylized representation of Missouri which is shown in Figure \ref{fig:minissouri}.
The state's total population is 180, and it is divided into three districts with population between 54 and 66.
Minissouri has two medium-sized cities and a substantial rural population with an overall partisan breakdown of 55.6% R to 44.4% D.
Voters' preferences are assumed to be stable across elections, and turnout is 100%.
In addition to partisan affiliation indicated by red or blue coloring, each voter also belongs to a racial group indicated by squares and circles.

```{r minissouri, fig.width=6.75, fig.pos='b'}
#| fig.cap: |
#|      Two potential districting plans for Minissouri,
#|      a Democratic gerrymander (a), and a close-to-proportional plan (b).
#|      Panel (c) shows which voters are harmed by the Democratic gerrymander,
#|      using the proportional plan (b) as a counterfactual.
#|      Panel (d) summarizes this across precincts, showing
#|      the average harm done to voters of each partisan and racial group.
include_graphics("figures/minissouri.pdf")
```

Now suppose the redistricting plan in Figure \ref{fig:minissouri}(a) has been enacted.
Despite the statewide popular vote favoring Republicans, under this plan Democrats will always win two of the three seats.
In contrast, a more proportional plan \ref{fig:minissouri}(b) elects two Republicans and one Democrat.

Intuitively, plan (a) unfairly packs Republicans into a rural district, depriving them of one seat.
To turn this intuition into something more concrete, we can identify which voters are harmed by the enactment of plan (a) over the alternative (b).
These voters are highlighted in Figure \ref{fig:minissouri}(c).
For example, in the upper leftmost precinct, there are three Republican square voters who are placed into a Democratic district in plan (a) but would have been placed into a Republican district in plan (b), and thus are harmed.
In contrast, Republican voters in the northeast city are not harmed, since they reside in a Democratic district in both the enacted and counterfactual plan.

@@

# Individual and Differential Harm {#sec:indiv-harm}

We formalize and generalize the concepts in Section \ref{sec:motivation}.
Let $\Voters$ be the set of individuals, or *voters*, in the region and time period.
Individuals vote for a candidate to represent districts, which are indexed by $\Distrs$.

@@

In an election $v$, the votes are tallied and the candidate with the most votes wins.
Using potential outcomes notation, we denote the winner of voter $i$'s district under districting plan $q$ and election $v$ as $\win_i(v,q)$.
This notation obscures the complex relationships and feedback effects between individuals, elections, and districting plans.
For example, different districting plans may lead to different candidates running, depending on district partisanship and incumbent residency.
The assumptions we make on these relationships will be made more explicit in Section \ref{sec:estimating}, where we discuss how to estimate our measures in practice.

## Harm {#subsec:harm}

The key intuition from Section \ref{sec:motivation} is that to be harmed, a voter must have a better outcome in the counterfactual world where a different districting plan was in place.
Letting $\tilde q$ denote a counterfactual districting plan, we define the *harm* to an individual $i$ in election $v$ held under districting plan $q$ as

@@

A second limitation is that $h_i(v,q,\tilde q)$ ignores any preferences a voter may have between the candidates that they did not vote for.
For example, a voter may consider candidates A and B nearly interchangeable, but may strongly dislike candidate C.
If the voter ultimately voted for candidate A, this harm measure makes no distinction between candidate B or candidate C winning the voter's district.
This reflects the simplistic nature of the cost function implicitly assumed in our proposed measure.
The problem could be remedied by replacing the indicator functions in the measure with other cost functions that consider ordinal preferences.
Ballot-level ranked-choice vote data, or candidate ideology scores, could be used to estimate such a cost function.
Here, for simplicity and to avoid making further assumptions, we focus on the indicator-based cost function.
Section \ref{sec:boston} further discusses the implications of this particular specification in the context of primary elections.

@@

The randomness in $V$ and $\tilde Q$ allows us to compare the enacted districting plan to many plausible counterfactual alternatives, and across a range of electoral environments.
At times we may wish to study on a particular election or a particular counterfactual plan.
Treating $V$ and $\tilde Q$ as random variables allows us to condition on particular values of each.
A particular specification of the distribution of $V$ and $\tilde Q$ is required only when *estimating* our harm measures in a particular application, a topic we return to in Section \ref{sec:estimating}.

@@

## Differential Harm and Fairness {#subsec:diff-harm}

@@

# Estimating the Harm Measures {#sec:estimating}

@@

Election results present a different set of modeling challenges, given the large number of factors that influence election outcomes and the general unpredictability of future events.
We limit our discussion here to the case of estimating harm and differential partisan harm in a two-party setting.
Estimating differential racial harm in the two-party setting can be done by first generating ecological inferences or survey estimates of party preference by race, and then applying methods analogous to those outlined here.
Multiparty or primary elections pose additional modeling challenges.
In Section \ref{sec:boston} we consider a spatial-racial model for primary elections with multiple candidates, to illustrate the application of our framework to these more complicated cases.

@@

# Connections to Current Practice for Analyzing Partisan and Racial Gerrymandering {#sec:connections}

@@

## Racial Vote Dilution {#subsec:racial}

@@

In both these cases, the proposed harm measures quantify the extent to which a challenged districting plan differentially harms minority voters.
They account for geographic sorting, political cohesion, and the individual preferences of every voter.
Sections \ref{sec:alabama} and \ref{sec:boston} demonstrate this approach in practice.

@@

The latter three metrics are grounded in intuition over how to reliably identify a partisan gerrymander, and not in the larger theoretical framework of partisan symmetry [@katz2020].
As such, they make no distinction between the quantity they are trying to estimate and a particular estimator or formula used in calculations.
This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section \ref{sec:estimating}.[^4]
This is an important conceptual distinction that is often lost in day-to-day applications of various gerrymandering measures.
Plugging in past election results without using an election model assumes too much about the stability of electoral dynamics and consequently can significantly understate the uncertainty in how a particular districting plan will perform into the future.

@@

Despite these differences, there are also empirical similarities in many cases.
The efficiency gap is often highly correlated with estimates of differential partisan harm under a stochastic uniform partisan swing model (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^6]
Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with estimates differential partisan harm as well.
Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in Section \ref{sec:nj}.

@@

# Application to Partisan Gerrymandering in New Jersey {#sec:nj}

@@

```{r nj-maps, fig.width=6.25}
#| fig.cap: |
#|      Baseline vote patterns (a), two hypothetical gerrymanders (b) and (c),
#|      and one randomly sampled plan (d).
#|      County lines are overlaid in white.
include_graphics("figures/nj_maps.pdf")
```

Figure \ref{fig:nj-maps} shows the baseline Democratic vote share by precinct, along with the Democratic and Republican proposals.[^8]
Using an average of the 2016 presidential and 2018 senatorial elections as a baseline, the Democratic proposal yields 9 Democratic seats and 3 Republican seats, while the Republican proposal has 7 Democratic and 5 Republican seats.

@@

We sample 10,000 districting plans across four independent runs of the SMC algorithm of @mccartan2020, as implemented in the R package `redist` [@redist], setting constraints to favor compact plans with district boundaries that tend to follow county lines.
One such plan is shown in Figure \ref{fig:nj-maps}(d).
For each of these sampled plans and the two gerrymanders, we estimate the average harm and differential partisan harm versus the simulated baseline,[^9] and calculate the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
The distribution of these measures is shown in Figure \ref{fig:nj-meas}.

@@

```{r nj-meas, fig.width=7}
#| fig.cap: |
#|      The distribution of the existing and proposed measures across 
#|      a sample of 10,000 plans,
#|      with the values for two gerrymanders overlaid as vertical lines.
#|      The dashed line indicates a value of zero.
include_graphics("figures/nj_meas.pdf")
```

@@

Figure \ref{fig:nj-pairs} in the appendix expands on Figure \ref{fig:nj-meas} to also show the correlations between these various measures.
Expected Democratic seats, the differential harm estimates, the efficiency gap, and declination are all moderately correlated with each other.
Similarly, partisan bias and the mean-median score are moderately correlated, but very weakly correlated with the expected number of Democratic seats.
Thus there exist many plans which give Democrats a higher-than-average share of the seats, but which have partisan bias and mean-median score favoring Republicans.
New Jersey is just one state, and while some of the correlations between measures here are found in other states, some are not.
Appendix \ref{app:corr} records these correlations across simulated plans in eight states.
The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which averages 0.79 in the eight states we study.
Seats are, of course, more discrete and cannot directly identify the local areas where gains and losses are derived.

# Application to Voting Rights Litigation in Alabama {#sec:alabama}

@@

As discussed in Section \ref{subsec:racial}, plaintiffs had to satisfy the first *Gingles* criterion by submitting one or more remedial maps demonstrating that a new Black opportunity-to-elect district could be drawn while meeting all other redistricting requirements.
In *Merrill,* the plaintiff's expert, Moon Duchin, submitted four such remedial plans, one of which is shown in Figure \ref{fig:al-maps}(e).
These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (Figure \ref{fig:al-maps}(d)), especially since the federal district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
In Appendix \ref{app:alsampling}, we repeat the analysis performed here on a different counterfactual consisting of plans sampled with a redistricting simulation algorithm; the results are broadly similar.

```{r al-maps, fig.width=6.75}
#| fig.cap: |
#|      Top row: baseline vote patterns (a),
#|      fraction of the voting-age population which is Black (b),
#|      the expected fraction of voters in each precinct which are harmed by 
#|      the 2020 enacted congressional districts versus the set of four remedial 
#|      plans (c).
#|      2020 enacted district boundaries are overlaid
#|      in black on the leftmost map and in white on the other maps.
#|      Bottom row: the challenged plan and a submitted remedial plan 
#|      (d) and (e) colored by each district's Black share of the voting-age
#|      population, and the expected number of harmed voters in each 
#|      race/partisan group, relative to the total number of such voters (f).
include_graphics("figures/al_maps.pdf")
```

@@

The model estimates, as expected, reveal that voting is highly polarized by race outside of the cities of Birmingham and Montgomery.
With these estimates in hand, we can calculate the average number of voters harmed in each precinct and racial group.
Figure \ref{fig:al-maps}(c) shows the estimated geographic distribution of harm.
Even without the ecological estimates, it is immediately clear, with reference to the partisan and racial patterns shown in Figures \ref{fig:al-maps}(a) and (b), that harm is concentrated among Black and Democratic voters in Birmingham, Montgomery, and the southern end of the Black Belt region.
These areas are excluded from the Black-majority district under the challenged plan, but are included in a Black-majority district in the average remedial plan.

This intuition is confirmed by a closer analysis, summarized in Figure \ref{fig:al-maps}(f).
Across the state, the estimated average harm is 7.9%, which translates to around 159,000 voters.
Of these harmed voters, around 123,000 are Black, while 36,000 are White.
Similarly, 137,000 are Democratic, while only 22,000 are Republican.
Of course, the total number of each of these groups of voters is very different.
Normalizing by group sizes, the expected harm for a Black Democrat is 22%, for a Black Republican is 3.7%, for a White Democrat is 8.3%, and for a White Republican is 1.7%.
Thus while the challenged plan differentially harms both Black voters and Democratic voters, it *especially* harms Black Democrats, who have an estimated differential harm of 13.8% versus White Democrats and of 18.3% versus Black Republicans.

@@

# Application to Primary Elections in Boston {#sec:boston}

@@

```{r boston-map, fig.width=6.75, fig.height=3.375}
#| fig.cap: |
#|      Racial composition of the city of Boston (left)
#|      and 2010 city council districts (right).
#|      Neighborhood boundaries are overlaid in black on the
#|      left map and in white on the right map.
include_graphics("figures/boston_map.pdf")
```

The city of Boston is governed by a mayor and a nonpartisan 13-member city council.
Nine of the councilors are elected from contiguous city council districts, and the remaining four councilors are elected at-large, with all councilors serving two-year terms.
Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^11]
This section will study the extent to which the current district boundaries differentially harm voters of different racial groups, in comparison to a completely race-neutral districting process.
Figure \ref{fig:boston-map} shows the racial composition of the city as well as the boundaries of the 2010 city council districts and the city's official neighborhoods.

@@

```{r boston-factors, fig.cap="", fig.width=7, fig.height=3.5}
#| fig.cap: |
#|      Candidate factor loadings, with approximate 80\% credible intervals shown as
#|      ellipses (left), and geographic variation in factor values by race (right).
#|      Neighborhood boundaries overlaid in black.
#|      Factor values in the Black second-dimension map are truncated so that variation in the other maps is visible.
include_graphics("figures/boston_factors.pdf")
```

@@

# Discussion {#sec:discussion}

@@

Figure \ref{fig:nj-pairs} expands on Figure \ref{fig:nj-meas} in the main text with pairwise scatter plots and correlation summaries.

>>>>

---
title: Individual and Differential Harm in Redistricting
date: last-modified
authors:
  - name: Cory McCartan
    affiliations:
      - name: Pennsylvania State University
        department: Department of Statistics
        address: 461 Pollock Road
        city: University Park
        region: PA
        postal-code: 16801
    email: mccartan@psu.edu
    url: https://corymccartan.com/
    corresponding: true
  - name: Christopher T. Kenny
    affiliations:
    - name: Harvard University
      department: Department of Government
      address: 1737 Cambridge Street
      city: Cambridge
      region: MA
      postal-code: "02138 "
    email: christopherkenny@fas.harvard.edu
    url: https://christophertkenny.com
abstract: |
  Social scientists have developed dozens of measures for assessing partisan bias in redistricting.
  But these measures are not easily adapted to other groups, 
  including groups defined by race, class, or geography.
  Nor are they applicable to single- or no-party contexts, such as local redistricting.
  To overcome these limitations, we propose a unified framework of harm for evaluating 
  the impacts of a districting plan on individual voters and the groups to which they belong. 
  We consider a voter harmed if their chosen candidate is not elected under the current plan, 
  but would be under a different plan.
  Harm improves on existing measures by both focusing on the choices of individual voters 
  and directly incorporating counterfactual plans.
  We discuss strategies for estimating harm, and demonstrate the utility of our framework 
  through analyses of partisan gerrymandering in New Jersey, voting rights 
  litigation in Alabama, and racial dynamics of Boston City Council elections.
thanks: |
  The authors thank Steve Ansolabehere, Matt Blackwell, Kosuke Imai, Gary King, 
  Shiro Kuriwaki, Tyler Simko, Nick Stephanopolous, and 
  participants of the Fall 2021 Harvard American Politics Research Workshop,
  Fall 2022 Harvard Applied Statistics Workshop, 
  and Imai Research Group for helpful comments and discussion.
  Replication code and data are available at <https://github.com/CoryMcCartan/harm-redistricting>.
keywords:
  - representation
  - redistricting
  - voting rights
  - individual harm
bibliography: references.bib
format: 
    cmc-article-pdf:
        fontsize: 11pt
        biblio-style: apsr
        include-in-header: "header.tex"
editor_options:
  markdown:
    wrap: sentence
---

@@

The harm framework combines two key ideas, which are motivated at greater length in Section \ref{sec-motivation}.
First, it operates at the level of the individual voter and examines whether each voter is represented by the candidate of their choice.
Doing so both unifies the approach to gerrymandering across contexts and also reflects a core goal of representative democracy: that elected officials represent the preferences of the voting public to the greatest extent possible [see also @brunell2010].

@@

The combination of individual voter choices and counterfactual districting plans defines our notion of *individual harm*: informally, a voter is harmed if they are more likely to be represented by the candidate of their choice under a different districting plan.
And by analyzing how harms fall differentially along partisan, racial, ideological, and other lines, we can connect individual harm to traditional notions of gerrymandering.
Section \ref{sec-motivation} illustrates this with a simple example, and the general definitions are formalized in Section \ref{sec-indiv-harm}.

A key feature of the proposed framework is its clear separation of theoretical quantities of interest and empirical strategies for estimating them [@katz2022]. This distinction is also missed by many critics of the efficiency gap [@veomett2018efficiency; @deford2020symmetry]. The challenge of estimating harm and differential harm from observed data is discussed in Section \ref{sec-estimating}. Recent advances in redistricting simulation, which can generate a large number of random districting plans that respect relevant legal constraints, are critical to this effort [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020].
These algorithmically-generated sets of districting plans can provide a realistic set of counterfactuals for our measures.

The harm framework has important similarities with the common-law approach to voting rights litigation.
To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker; @issakarl98]. This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidate [@2004vieth; @2018league; @2019lewis].
Indeed, a primary factor in the Supreme Court's ruling that partisan gerrymandering is not justiciable was that existing redistricting measures are not clearly tied to individual injury [@2017gill].
A more detailed discussion of the connections between the proposed framework and existing practices for analyzing racial and partisan gerrymandering may be found in Section \ref{sec-connections}.

Harm and differential harm are more than a useful conceptual unification---they are practically useful in a wide variety of redistricting analyses.
A brief study of partisan gerrymandering of New Jersey congressional districts in Section \ref{sec-nj} shows how existing gerrymandering metrics can make incorrect conclusions about the magnitude and even the direction of partisan gerrymanders.
Section \ref{sec-alabama}, studies voting rights litigation for Alabama congressional districts, and uses differential harm to separate the racial and partisan effects of the enacted districting plan.
Section \ref{sec-boston} demonstrates the application of our framework to local redistricting, and quantifies the benefits of race-aware city council district boundaries to Black voters in the city of Boston.
Section \ref{sec-discussion} concludes.

# Motivation {#sec-motivation}

The core idea motivating individual and differential harm is a simple one.
Consider *Minissouri*, a stylized representation of Missouri which is shown in Figure \ref{fig-minissouri}.
The state's total population is 180, and it is divided into three districts with population between 54 and 66.
Minissouri has two medium-sized cities and a substantial rural population with an overall partisan breakdown of 55.6% R to 44.4% D.
Voters' preferences are assumed to be stable across elections, and turnout is 100%.
In addition to partisan affiliation indicated by red or blue coloring, each voter also belongs to a racial group indicated by squares and circles.

```{r fig-minissouri, fig.width=6.75}
#| fig.cap: |
#|      Two potential districting plans for Minissouri,
#|      a Democratic gerrymander (a), and a close-to-proportional plan (b).
#|      Panel (c) shows which voters are harmed by the Democratic gerrymander,
#|      using the proportional plan (b) as a counterfactual.
#|      Panel (d) summarizes this across precincts, showing
#|      the average harm done to voters of each partisan and racial group.
include_graphics("figures/minissouri.pdf")
```

Now suppose the redistricting plan in Figure \ref{fig-minissouri}(a) has been enacted.
Despite the statewide popular vote favoring Republicans, under this plan Democrats will always win two of the three seats.
In contrast, a more proportional plan \ref{fig-minissouri}(b) elects two Republicans and one Democrat.

Intuitively, plan (a) unfairly packs Republicans into a rural district, depriving them of one seat.
To turn this intuition into something more concrete, we can identify which voters are harmed by the enactment of plan (a) over the alternative (b).
These voters are highlighted in Figure \ref{fig-minissouri}(c).
For example, in the upper leftmost precinct, there are three Republican square voters who are placed into a Democratic district in plan (a) but would have been placed into a Republican district in plan (b), and thus are harmed.
In contrast, Republican voters in the northeast city are not harmed, since they reside in a Democratic district in both the enacted and counterfactual plan.

@@

# Individual and Differential Harm {#sec-indiv-harm}

We formalize and generalize the concepts in Section \ref{sec-motivation}.
Let $\Voters$ be the set of individuals, or *voters*, in the region and time period.
Individuals vote for a candidate to represent districts, which are indexed by $\Distrs$.

@@

In an election $v$, the votes are tallied and the candidate with the most votes wins.
Using potential outcomes notation, we denote the winner of voter $i$'s district under districting plan $q$ and election $v$ as $\win_i(v,q)$.
This notation obscures the complex relationships and feedback effects between individuals, elections, and districting plans.
For example, different districting plans may lead to different candidates running, depending on district partisanship and incumbent residency.
The assumptions we make on these relationships will be made more explicit in Section \ref{sec-estimating}, where we discuss how to estimate our measures in practice.

## Harm {#subsec-harm}

The key intuition from Section \ref{sec-motivation} is that to be harmed, a voter must have a better outcome in the counterfactual world where a different districting plan was in place.
Letting $\tilde q$ denote a counterfactual districting plan, we define the *harm* to an individual $i$ in election $v$ held under districting plan $q$ as

@@

A second limitation is that $h_i(v,q,\tilde q)$ ignores any preferences a voter may have between the candidates that they did not vote for.
For example, a voter may consider candidates A and B nearly interchangeable, but may strongly dislike candidate C.
If the voter ultimately voted for candidate A, this harm measure makes no distinction between candidate B or candidate C winning the voter's district.
This reflects the simplistic nature of the cost function implicitly assumed in our proposed measure.
The problem could be remedied by replacing the indicator functions in the measure with other cost functions that consider ordinal preferences.
Ballot-level ranked-choice vote data, or candidate ideology scores, could be used to estimate such a cost function.
Here, for simplicity and to avoid making further assumptions, we focus on the indicator-based cost function.
Section \ref{sec-boston} further discusses the implications of this particular specification in the context of primary elections.

@@

The randomness in $V$ and $\tilde Q$ allows us to compare the enacted districting plan to many plausible counterfactual alternatives, and across a range of electoral environments.
At times we may wish to study on a particular election or a particular counterfactual plan.
Treating $V$ and $\tilde Q$ as random variables allows us to condition on particular values of each.
A particular specification of the distribution of $V$ and $\tilde Q$ is required only when *estimating* our harm measures in a particular application, a topic we return to in Section \ref{sec-estimating}.

@@

## Differential Harm and Fairness {#subsec-diff-harm}

@@

# Estimating the Harm Measures {#sec-estimating}

@@

Election results present a different set of modeling challenges, given the large number of factors that influence election outcomes and the general unpredictability of future events.
We limit our discussion here to the case of estimating harm and differential partisan harm in a two-party setting.
Estimating differential racial harm in the two-party setting can be done by first generating ecological inferences or survey estimates of party preference by race, and then applying methods analogous to those outlined here.
Multiparty or primary elections pose additional modeling challenges.
In Section \ref{sec-boston} we consider a spatial-racial model for primary elections with multiple candidates, to illustrate the application of our framework to these more complicated cases.

@@

# Connections to Current Practice for Analyzing Partisan and Racial Gerrymandering {#sec-connections}

@@

## Racial Vote Dilution {#subsec-racial}

@@

In both these cases, the proposed harm measures quantify the extent to which a challenged districting plan differentially harms minority voters.
They account for geographic sorting, political cohesion, and the individual preferences of every voter.
Sections \ref{sec-alabama} and \ref{sec-boston} demonstrate this approach in practice.

@@

The latter three metrics are grounded in intuition over how to reliably identify a partisan gerrymander, and not in the larger theoretical framework of partisan symmetry [@katz2020].
As such, they make no distinction between the quantity they are trying to estimate and a particular estimator or formula used in calculations.
This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section \ref{sec-estimating}.[^4]
This is an important conceptual distinction that is often lost in day-to-day applications of various gerrymandering measures.
Plugging in past election results without using an election model assumes too much about the stability of electoral dynamics and consequently can significantly understate the uncertainty in how a particular districting plan will perform into the future.

@@

Despite these differences, there are also empirical similarities in many cases.
The efficiency gap is often highly correlated with estimates of differential partisan harm under a stochastic uniform partisan swing model (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^6]
Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with estimates differential partisan harm as well.
Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in Section \ref{sec-nj}.

@@

# Application to Partisan Gerrymandering in New Jersey {#sec-nj}

@@

```{r fig-nj-maps, fig.width=6.25}
#| fig.cap: |
#|      Baseline vote patterns (a), two hypothetical gerrymanders (b) and (c),
#|      and one randomly sampled plan (d).
#|      County lines are overlaid in white.
include_graphics("figures/nj_maps.pdf")
```

Figure \ref{fig-nj-maps} shows the baseline Democratic vote share by precinct, along with the Democratic and Republican proposals.[^8]
Using an average of the 2016 presidential and 2018 senatorial elections as a baseline, the Democratic proposal yields 9 Democratic seats and 3 Republican seats, while the Republican proposal has 7 Democratic and 5 Republican seats.

@@

We sample 10,000 districting plans across four independent runs of the SMC algorithm of @mccartan2020, as implemented in the R package `redist` [@redist], setting constraints to favor compact plans with district boundaries that tend to follow county lines.
One such plan is shown in Figure \ref{fig-nj-maps}(d).
For each of these sampled plans and the two gerrymanders, we estimate the average harm and differential partisan harm versus the simulated baseline,[^9] and calculate the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
The distribution of these measures is shown in Figure \ref{fig-nj-meas}.

@@

```{r fig-nj-meas, fig.width=7}
#| fig.cap: |
#|      The distribution of the existing and proposed measures across 
#|      a sample of 10,000 plans,
#|      with the values for two gerrymanders overlaid as vertical lines.
#|      The dashed line indicates a value of zero.
include_graphics("figures/nj_meas.pdf")
```

@@

Figure \ref{fig-nj-pairs} in the appendix expands on Figure \ref{fig-nj-meas} to also show the correlations between these various measures.
Expected Democratic seats, the differential harm estimates, the efficiency gap, and declination are all moderately correlated with each other.
Similarly, partisan bias and the mean-median score are moderately correlated, but very weakly correlated with the expected number of Democratic seats.
Thus there exist many plans which give Democrats a higher-than-average share of the seats, but which have partisan bias and mean-median score favoring Republicans.
New Jersey is just one state, and while some of the correlations between measures here are found in other states, some are not.
Appendix \ref{app:corr} records these correlations across simulated plans in eight states.
The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which averages 0.79 in the eight states we study.
Seats are, of course, more discrete and cannot directly identify the local areas where gains and losses are derived.

# Application to Voting Rights Litigation in Alabama {#sec-alabama}

@@

As discussed in Section \ref{subsec-racial}, plaintiffs had to satisfy the first *Gingles* criterion by submitting one or more remedial maps demonstrating that a new Black opportunity-to-elect district could be drawn while meeting all other redistricting requirements.
In *Merrill,* the plaintiff's expert, Moon Duchin, submitted four such remedial plans, one of which is shown in Figure \ref{fig-al-maps}(e).
These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (Figure \ref{fig-al-maps}(d)), especially since the federal district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
In Appendix \ref{app:alsampling}, we repeat the analysis performed here on a different counterfactual consisting of plans sampled with a redistricting simulation algorithm; the results are broadly similar.

```{r fig-al-maps, fig.width=6.75}
#| fig.cap: |
#|      Top row: baseline vote patterns (a),
#|      fraction of the voting-age population which is Black (b),
#|      the expected fraction of voters in each precinct which are harmed by 
#|      the 2020 enacted congressional districts versus the set of four remedial 
#|      plans (c).
#|      2020 enacted district boundaries are overlaid
#|      in black on the leftmost map and in white on the other maps.
#|      Bottom row: the challenged plan and a submitted remedial plan 
#|      (d) and (e) colored by each district's Black share of the voting-age
#|      population, and the expected number of harmed voters in each 
#|      race/partisan group, relative to the total number of such voters (f).
include_graphics("figures/al_maps.pdf")
```

@@

The model estimates, as expected, reveal that voting is highly polarized by race outside of the cities of Birmingham and Montgomery.
With these estimates in hand, we can calculate the average number of voters harmed in each precinct and racial group.
Figure \ref{fig-al-maps}(c) shows the estimated geographic distribution of harm.
Even without the ecological estimates, it is immediately clear, with reference to the partisan and racial patterns shown in Figures \ref{fig-al-maps}(a) and (b), that harm is concentrated among Black and Democratic voters in Birmingham, Montgomery, and the southern end of the Black Belt region.
These areas are excluded from the Black-majority district under the challenged plan, but are included in a Black-majority district in the average remedial plan.

This intuition is confirmed by a closer analysis, summarized in Figure \ref{fig-al-maps}(f).
Across the state, the estimated average harm is 7.9%, which translates to around 159,000 voters.
Of these harmed voters, around 123,000 are Black, while 36,000 are White.
Similarly, 137,000 are Democratic, while only 22,000 are Republican.
Of course, the total number of each of these groups of voters is very different.
Normalizing by group sizes, the expected harm for a Black Democrat is 22%, for a Black Republican is 3.7%, for a White Democrat is 8.3%, and for a White Republican is 1.7%.
Thus while the challenged plan differentially harms both Black voters and Democratic voters, it *especially* harms Black Democrats, who have an estimated differential harm of 13.8% versus White Democrats and of 18.3% versus Black Republicans.

@@

# Application to Primary Elections in Boston {#sec-boston}

@@

```{r fig-boston-map, fig.width=6.75, fig.height=3.375}
#| fig.cap: |
#|      Racial composition of the city of Boston (left)
#|      and 2010 city council districts (right).
#|      Neighborhood boundaries are overlaid in black on the
#|      left map and in white on the right map.
include_graphics("figures/boston_map.pdf")
```

The city of Boston is governed by a mayor and a nonpartisan 13-member city council.
Nine of the councilors are elected from contiguous city council districts, and the remaining four councilors are elected at-large, with all councilors serving two-year terms.
Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^11]
This section will study the extent to which the current district boundaries differentially harm voters of different racial groups, in comparison to a completely race-neutral districting process.
Figure \ref{fig-boston-map} shows the racial composition of the city as well as the boundaries of the 2010 city council districts and the city's official neighborhoods.

@@

```{r fig-boston-factors, fig.cap="", fig.width=7, fig.height=3.5}
#| fig.cap: |
#|      Candidate factor loadings, with approximate 80\% credible intervals shown as
#|      ellipses (left), and geographic variation in factor values by race (right).
#|      Neighborhood boundaries overlaid in black.
#|      Factor values in the Black second-dimension map are truncated so that variation in the other maps is visible.
include_graphics("figures/boston_factors.pdf")
```

@@

# Discussion {#sec-discussion}

@@

Figure \ref{fig-nj-pairs} expands on Figure \ref{fig-nj-meas} in the main text with pairwise scatter plots and correlation summaries.
